% Manuscript source — Astronomy and Computing (Elsevier).
\documentclass[preprint,12pt]{elsarticle}

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[english]{babel}
\usepackage[protrusion=true,expansion=false]{microtype}

\setlength{\emergencystretch}{2em}
\hyphenpenalty=800
\exhyphenpenalty=800
\journal{Astronomy and Computing}

\begin{document}

\begin{frontmatter}

\title{A Reproducible Control-Validation Workflow for Circular-Shift and
Surrogate Tests of Galactic Radial Profiles}


\author[aff1]{Vitor M. F. Figueiredo\corref{cor1}}
\ead[url]{https://orcid.org/0009-0004-7358-4622}
\cortext[cor1]{Corresponding author.}
\address[aff1]{Independent Researcher, Nisa, Portugal}

\begin{abstract}
\begin{sloppypar}
Correlations between radially organized covariates and the residuals of
galaxy rotation-curve fits are used as diagnostics of inner-region
structure. Assessing such correlations requires controls that account for
the strong radial organization of the underlying data, because a
correlation can exceed a simple label-shuffle comparison while reflecting
only a broad, smooth radial trend. We present a reproducible,
branch-separated control-validation workflow for radial-structure
diagnostics in SPARC-like rotation-curve data. For a partial-rank branch
of 134 galaxies, a radius-conditioned partial-rank statistic is compared
against stochastic shuffle/permutation empirical control distributions and
against exact circular-shift randomization distributions that preserve
radial ordering. The observed absolute partial-rank statistic has
across-galaxy median $\simeq 0.509$; the shuffle control
median-of-medians is $\simeq 0.182$, whereas the ordering-preserving exact
circular-shift control median-of-medians is $\simeq 0.532$, comparable to
the observed statistic. For a separate spectral-control branch of 76
galaxies, an amplitude-based spectral-concentration statistic is evaluated
against amplitude-preserving random-phase controls; because the statistic
is amplitude-based, equality with these controls holds by construction and
is reported as a caveat rather than as evidence. Fixed smooth profiles are
included as deterministic comparators, not stochastic controls. We
document the workflow, its deterministic-seed implementation, and an
audited descriptive demonstration for reuse, and make no physical
interpretation or model-validation claim.
\end{sloppypar}
\end{abstract}

\begin{keyword}
methods: statistical \sep methods: data analysis \sep galaxies: kinematics
and dynamics \sep surrogate data \sep permutation testing \sep
reproducibility
\end{keyword}

\end{frontmatter}

\input{sections/introduction}
\input{sections/data}
\input{sections/methods}
\input{sections/demonstration}
% Results redundancy resolved: the descriptive results live in the
% Demonstration section. The former empty Results stub is intentionally
% not included (see notes/FULL_PAPER_MAJOR_REVISION_REPORT.md, Objective 3).
\input{sections/discussion}
\input{sections/reproducibility}
\input{sections/conclusion}

% Elsevier requirement: this declaration must appear immediately above the
% references.
\section*{Declaration of generative AI and AI-assisted technologies in the
writing process}
\begin{sloppypar}
During the preparation of this work, the author used OpenAI ChatGPT and Codex, Anthropic Claude and claude code, Google Gemini and Grok for drafting assistance, language refinement, code-review support, and the preparation of audit checklists. After using these tools, the author reviewed and edited the content as needed and takes full responsibility for the content of the publication.
\end{sloppypar}


\bibliographystyle{elsarticle-harv}
\bibliography{references}

\end{document}
