
\section{Methods}
\label{sec:methods}

\subsection{Diagnostic definitions}
\label{sec:diagnostics}

For galaxy $g$, let $i=1,\ldots,n_g$ index its frozen valid radial bins. Let
$r_{gi}$ denote the radius-conditioning coordinate,
$y_{gi}=|\log_{10}(v_{\mathrm{rec},gi}/v_{\mathrm{obs},gi})|$ the residual
magnitude, and $\sigma_{gi}$ the reconstructed total radial covariate.

\subsubsection{Radial-template covariate and partial-rank statistic}

The normalized covariate and radial-template function are
\begin{equation}
c_{gi} =
\frac{\sigma_{gi}}
{\operatorname{median}\{\sigma_{gj}:\sigma_{gj}>0,\ \sigma_{gj}\ {\rm finite}\}},
\end{equation}
and
\begin{equation}
t(c)=\operatorname{clip}_{[0,1]}
\left[
\exp\left\{-\left(\frac{\log_{10}(\max(c,10^{-12}))}{0.50}\right)^2\right\}
\right].
\end{equation}
A scale fallback of one is used when no finite positive covariate value
exists, and nonfinite normalized entries are replaced by zero.

Let $\operatorname{rank}(\cdot)$ assign average ranks to ties and
$X_g=[\mathbf{1},\operatorname{rank}(r_g)]$. For any vector $z_g$, define
\begin{equation}
\widehat{\beta}_z=\underset{\beta}{\arg\min}\,
\|\operatorname{rank}(z_g)-X_g\beta\|_2^2,\qquad
u_z=\operatorname{rank}(z_g)-X_g\widehat{\beta}_z,
\end{equation}
implemented through least-squares fitting. The signed statistic is
\begin{equation}
\rho_g=\operatorname{corr}\left(u_{t(c)},u_y\right),
\end{equation}
and the production-facing headline summary uses $|\rho_g|$.

\subsubsection{Spectral concentration}

Each valid profile is linearly interpolated onto 64 evenly spaced points in
$\log_{10}(R/\operatorname{median}_{+}(R))$ and mean-centered. After a
real-valued Fourier transform and removal of the DC component, define
\begin{equation}
p_k=\frac{|F_k|^2}{\sum_j |F_j|^2},
\qquad
C=1-\frac{-\sum_k p_k\log p_k}{\log K},
\end{equation}
where $K$ is the number of retained non-DC power bins.

\subsection{Control families and empirical summaries}
\label{sec:controls}

Permutation and surrogate procedures are established tools in general
statistical and correlation-testing contexts
\citep{SchreiberSchmitz2000,PhipsonSmyth2010,YuHutson2022}.

\subsubsection{Shuffle/permutation controls}

For each partial-branch galaxy, $M=1000$ deterministic-seed realizations
permute only the normalized radial covariate. Radius and residual vectors
remain frozen, and the template function is applied after permutation. This
general baseline does not imply common practice in rotation-curve analysis.

\subsubsection{Exact circular-shift randomization distributions}

For a galaxy with $n_g$ valid bins, the production-facing
ordering-preserving control uses every offset in
\begin{equation}
\mathcal{S}_g=\{1,2,\ldots,n_g-1\}
\end{equation}
exactly once. Offset zero is excluded. For each $s\in\mathcal{S}_g$,
$c_g^{(s)}=\operatorname{roll}(c_g,s)$ is transformed and evaluated with the
same radius-conditioned statistic.

Cyclic wrap-around is an operational finite-sample randomization convention,
not a claim that the innermost and outermost radial bins are physically
adjacent. Only the normalized radial covariate is rolled; the radius vector,
residual profile, finite sample size, and frozen validity mask remain fixed.
Applying every non-identity offset preserves the ordered covariate sequence
under translation and defines an exact finite randomization space over all
nontrivial cyclic alignments. The comparison tests sensitivity to radial
ordering and alignment between the radial-template covariate and residual
profile, conditional on this operational convention.

\subsubsection{Amplitude-preserving random-phase controls}

For each spectral-valid residual and radial-template profile, $M=1000$
controls preserve Fourier amplitudes while randomizing the unconstrained
phases at positive frequencies. The mean-centered DC component is zero; the
even-length Nyquist coefficient remains real with a randomized sign; and an
inverse real-valued Fourier transform restores the profile
\citep{Theiler1992,SchreiberSchmitz2000}. More constrained surrogate families
can preserve additional data properties \citep{SchreiberSchmitz1996}, but the
present random-phase tier is intentionally limited to amplitude preservation
and is interpreted only within that control-family definition.

The random-phase comparison is included as a sensitivity diagnostic for the
spectral statistic: because spectral concentration is amplitude-based,
amplitude-preserving phase randomization demonstrates which quantities are
invariant by construction rather than providing independent physical
evidence.

\subsubsection{Smooth fixed references}

On a 64-point coordinate $x\in[0,1]$, the fixed profiles are
\begin{equation}
\exp(-x),\qquad \max(1-x,0),\qquad \max(1-x,0)^2.
\end{equation}
They are fixed references, not stochastic controls, and receive no empirical
percentile or p-value.

\subsubsection{Empirical summaries}

For finite observed value $z_{\rm obs}$, finite controls
$\{z_m\}_{m=1}^{M_v}$, and $\epsilon=10^{-12}$, define
\begin{equation}
L=\#\{z_m<z_{\rm obs}-\epsilon\},\quad
E=\#\{|z_m-z_{\rm obs}|\leq\epsilon\},\quad
G=\#\{z_m>z_{\rm obs}+\epsilon\},
\end{equation}
with $L+E+G=M_v$. The bounded mid-rank percentile is
\begin{equation}
q=\operatorname{clip}_{[0,1]}\left(\frac{L+E/2}{M_v}\right).
\end{equation}
For signed or absolute random-tier statistics,
\begin{equation}
p_{\rm MC}=\operatorname{clip}_{[0,1]}
\left[
\frac{1+\#\{|z_m|\geq |z_{\rm obs}|-\epsilon\}}{M_v+1}
\right].
\end{equation}
The same formula is used for the exhaustive nontrivial circular-shift set.
For the nonnegative phase-blind spectral statistic,
$p_{\rm spectral}=\min\{1,2\min(q,1-q)\}$. Fixed references receive neither
an empirical percentile nor an empirical p-value.

